---
title: "Week 2 Assignment"
author: "Jessalyn Chuang"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary packages.
library(here)
library(tidyverse)
library(viridis)
```

# (1) Measures of location and data spread

```{r}
# Load necessary data.
crab_data <- read_csv(file=here("Labs/week2/pie_crab.csv"))
mean(crab_data$size)
median(crab_data$size)
sd(crab_data$size)
```


##a. 
What is the mean fiddler crab carapace size?
The mean fiddler crab carapace size is 14.65801 mm.

##b. 
What is the median fiddler crab carapace size?
The median fiddler crab carapace size is 14.65801 mm.

##c. 
What is the standard deviation in fiddler crab carapace size?
The standard deviation in fiddler crab carapace size is 3.503304 mm.

# (2) Creating Histograms

##a. 
One of the ways to visualize data that we have discussed in class is using histograms. 
Create a figure that is a histogram of fiddler crab carapace size (across all sites), and take care to make
sure your figure uses an appropriate number of bins and includes a caption. (Hint: Learn how
to use the geom_histogram() call to achieve this by typing ?geom_histogram() into your console
or Google.)

```{r}
ggplot(data = crab_data, aes(x = size)) +
  geom_histogram(bins = 15, fill = "lightpink", color = "black") + 
  labs(x = "Size (mm)",
       y = "Count",
       title = "Histogram of Fiddler Crab Carapace Size Across all Sites)")
```

##b. 
Create a second figure that now displays separate histograms for fiddler crab carapace size
for each of the 13 sites. Again, make sure axes are appropriately labeled, and your figure
includes a caption. (Hint: Use the facet_wrap() call and explore its functionality either in the
console or via Google.)

```{r}
ggplot(data = crab_data, aes(x = size)) +
  geom_histogram(bins = 15, fill = "pink", color = "black") + 
  facet_wrap(~ site) +  # Create separate histograms for each site
  labs(x = "Size (mm)",
       y = "Count",
       title = "Histogram of Fiddler Crab Carapace Size by Site",
       caption = "Each panel represents a separate site, showing the distribution of carapace sizes.") 
```

##c. 
How do the raw values of fiddler crab carapace size in your first figure compare to the mean
and median values you calculated above? Based on your second figure, do any sites display
strong diﬀerences in their data distributions?

The mean and the median values fall into the section of the histogram with the highest count.

Yes, certain sites exhibit strong differences in their data distributions. Here are key observations:

Sites with Distinct Distributions:
GTM and PIE:

These sites show a strong peak at smaller carapace sizes (~10-15mm), with a right-skewed distribution.
They have a clear concentration of smaller crabs compared to other sites.
CT and DB:

These sites exhibit a more symmetric, bell-shaped distribution, suggesting a more even spread of carapace sizes.
Their distributions resemble a normal distribution.
ZI and SI:

Both sites have left-skewed distributions, with a higher frequency of smaller sizes.
The presence of a strong peak at lower sizes (~10mm) differentiates them from other sites.
RC and NB:

These sites have a broader range of sizes but appear to have bimodal distributions (two peaks), suggesting the presence of multiple size groups.
Sites with Similar Distributions:
Sites such as NIB, VCR, and JC show distributions that are more evenly spread, lacking strong skewness or multimodal patterns.
Their size distributions appear balanced across a range of carapace sizes.
Conclusion:
GTM, PIE, ZI, and SI show strong differences, likely due to environmental or biological factors influencing crab growth.
CT and DB have a symmetrical distribution.
RC and NB might have bimodal distributions, suggesting possible subpopulations or size classes.

# (3) Data Visualization - the Golden Rules

##a. 
Make a new plot examining any other relationship you are interested in visualizing using

the dataset. It can be a scatterplot (geom_point()), a jitter plot (geom_jitter()), a boxplot
(geom_boxplot()), a histogram (geom_histogram()), or any other data display you’d like. Take
care to follow each of the golden rules of data visualization - Accuracy, Clarity, Aesthestics -
and be sure to include a descriptive figure caption.

##b. 
Now, using the same base figure you created above in part a, create the worst version of

that same figure that you possibly can. Think color, font size, replacing points with emojis,
etc. - the uglier and more confusing the better! You must also include a figure caption, but
it, too, should be unhelpful.