Setup

Create a new project for this week’s lab named week6 in the ENV710 folder. Open and save a new R Markdown
script named week6script.Rmd in this project. Edit the title of the script accordingly, and add a first Setup
header to your script. Then, create and run a new code chunk to load in the packages you will use today.

```{r}
# Load necessary packages.
library(here)
library(tidyverse)
library(moments)
library(palmerpenguins)

# Save the palmerpenguins dataset into your Environment.
penguin_dat <- penguins
```

Functions for grouping and summarizing data

When you’d like to quickly summarize the contents of a dataframe and impose a particular grouping on the
data, the group_by() and summarize() functions allow you to do exactly that.

The group_by() function takes the name of the column that you want to group by, which needs to be
categorical. When using group_by(), however, don’t forget to ungroup() your data the close of your pipe so
that grouping is not carried forward into other analyses you might want to conduct.

The summarize() function will perform the summary operations on the groups you’ve specified and created
a new dataframe containing the summary outputs.

```{r}
# Calculate mean and standard deviation values of penguin flipper
# length according to the island they live on.
flipper_summary <- penguin_dat %>%
  group_by(island) %>%
  summarize(mean = mean(flipper_length_mm, na.rm = TRUE),
  sd = sd(flipper_length_mm, na.rm = TRUE)) %>%
ungroup()
```

Practice Problem: Create a new dataframe that calculates the median values and variance in bill length
according to penguin sex. Don’t forget to ungroup your data at the end of your pipe!

```{r}
bill_summary <- penguin_dat %>% 
  group_by(sex) %>% 
  summarize(median = median(bill_length_mm, na.rm = TRUE),
            variance = var(bill_length_mm, na.rm = TRUE)) %>% 
  ungroup()
```

Conducting Analysis of Variance (ANOVA) tests

Analysis of Variance or ANOVA tests may be used when you want to examine diﬀerences in more than two
sample means. Remember the following hypotheses apply to an ANOVA:

Null Hypothesis: All sample means are equal.
Alternative Hypothesis: At least two sample means, among all samples, are not equal.

Before running an ANOVA, you must also ensure that your data meets the following requirements:
• Samples are independent of one another.
• Data within samples is normally distributed.
• Variances among samples are equal.

Create a new header in your script titled One-Way ANOVAs.

#ANOVAs

Examining Data Distribution
We will first examine the data we are using to ensure that sample distributions are approximately normal.

```{r}
# Examine penguin body mass by species
penguin_fig1 <- ggplot(penguin_dat, aes(x = body_mass_g,
fill = species)) + # base plot
geom_histogram() + # creates histograms
scale_fill_manual(values = c("darkorange","purple","cyan4")) + # customize colors
labs(x = "Body Mass (g)",
y = "Count") + # edits labels
facet_grid(.~species) + # facets by species to avoid overlap
theme_bw() + # removes the grey background
theme(legend.position = "none") # removes the legend
# View figure.
penguin_fig1
```

All but Gentoo appear reliably normally distributed, so will also calculate skew and kurtosis for this species
to be more certain of moving forward with the one-way ANOVA test.

```{r}
# Filter out only Gentoo values.
gentoo_dat <- penguin_dat %>%
filter(species== "Gentoo")
# Calculate skew.
skewness(gentoo_dat$body_mass_g, na.rm = TRUE)
# Calculate kurtosis.
kurtosis(gentoo_dat$body_mass_g, na.rm = TRUE)
```



The data appears to be approximately symmetrical (skewness = 0.06) and moderately platykurtic (kurtosis
= 2.26), so we will proceed with the assumption that these body mass data for the Gentoo species are
approximately normally distributed.

Note, many data you will deal with in the future will not be normally distributed. In these cases
you have two options on how to proceed:

(1) Apply a log transformation. This is standard practice to adjust raw values so that they do conform to
a normal distribution and you may continue with applying a statistical test that requires normally-
distributed data, like a one-way ANOVA. In R, you may apply either a log based 10 transformation
(log10()) or a natural log transformation (log()) - either option is frequently used, just be sure to
state clearly which you have performed.

(2) Stop and re-consider what kind of statistical test you should use. Later this week and early next, we’ll
cover a few alternative options when you encounter data that is not normally distributed but you would
still like to compare values between groups.

##Bartlett’s Test for Equal Variance
The next thing we need to ensure is that variances are approximately equal across species. The Bartlett test
for homogeneity of variances is based on a null hypothesis that the variance between all groups included in
the data are the same. Typically, you will use a 95% confidence level (5% significant level) when applying
this test.

```{r}
# Perform Bartlett test.
penguin_var <- bartlett.test(penguin_dat$body_mass_g, penguin_dat$species)
# Examine results.
penguin_var
```

Because p= 0.05, we do not have suﬃcient evidence to reject the null hypothesis, so we may assume that
variance across the three groups (species) in this dataset are equal and proceed with conducting a one-way
ANOVA.

There is also the helpful rule of thumb that states that if the largest sample variance is less than 4 times the
smallest sample variance, you may still assume variances are equal across samples and conduct an ANOVA.

##Analysis of Variance (ANOVA)

```{r}
# Perform a one-way ANOVA.
penguin_ANOVA <- aov(body_mass_g~ species, data = penguin_dat)
# Examine the results.
summary(penguin_ANOVA)
```

If p ≥ 0.05, you may conclude that there are no significant diﬀerences between mean values of any of the
samples included in your dataset. If this is the case, you may stop here and report your results as such.
If p < 0.05, you may conclude that there is a significant diﬀerence in mean values between at least two of
your samples and continue forward to conduct a Tukey’s Honestly Significant Diﬀerence (HSD) test.

#Post-hoc Tukey’s HSD Test

```{r}
# Perform a Tukey's HSD test.
penguin_Tukey <- TukeyHSD(penguin_ANOVA)
# Examine the results.
penguin_Tukey
```

Notice how the results of the Tukey’s HSD display the p-value for each of the pair-wise comparisons between
species. Based on these values, we may conclude that there is in fact a significant diﬀerence in body mass
between the Gentoo and Adelie species as well as the Gentoo and Chinstrap species.

#Communicating Results

When graphically presenting the results of an ANOVA, it is common to include small letters above each of
the samples to notate which samples were significant from others. We’ll do exactly that in the following plot.
We’ll first calculate the mean and standard deviation values and then overlay those over top of the raw data
points to provide better context for data distribution amongst the three species.

```{r}
# Calculate means and standard deviation.
penguin_summary <- penguin_dat %>%
group_by(species) %>%
summarize(mean = mean(body_mass_g, na.rm = TRUE),
sd = sd(body_mass_g, na.rm = TRUE)) %>%
ungroup()
# We also need to rename the "mean" column - you
# will see why in a moment.
penguin_summary <- penguin_summary %>%
rename(body_mass_g = mean)

# Create a plot to display results.
# Because we are using multiple datasets in the same figure,
# we will specify the aesthetics in each individual data call
penguin_fig2 <- ggplot() +
  # add raw data points
  geom_jitter(data = penguin_dat, aes(x = species, y = body_mass_g, color = species), alpha = 0.5, size = 0.5) +
  # add summary statistics
  geom_point(data = penguin_summary, aes(x = species, y = body_mass_g, color = species), size = 3) +
  # add error bars 
  geom_errorbar(data = penguin_summary,aes(x = species, ymin = body_mass_g-sd, ymax = body_mass_g+sd, color = species), width = 0.10, size = 1) +
  # add text annotations
  annotate("text", x = "Adelie", y = 5500, label = "a", size = 8) +
  annotate("text", x = "Chinstrap", y = 5500, label = "a", size = 8) + annotate("text", x = "Gentoo", y = 6100,
label = "b", size = 8) +
  # edit colors
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  # label axes
  labs(x = "Species", y = "Body Mass (g)") +
  # remove grey background
  theme_bw() +
  # remove legend
  theme(legend.position = "none")
# View figure.
penguin_fig2
```

When reporting the results of an ANOVA it’s best practice to include many of the measures of central
tendency and data spread as we’ve seen in previous weeks. Furthermore, you’ll want to include the degrees of
freedom, the F-statistic, and p-value of the ANOVA as well as the p-values for any of the post-hoc tests that
yielded significant diﬀerences. An example of how we might report our findings from this particular test can
be found below:

Penguin species displayed significant diﬀerences in body mass as determined by one-way ANOVA (F(2, 339)
= 343.6, p < 0.001). Post-hoc testing by Tukey’s HSD revealed that mean body mass for Adelie (mean =
3,700 g, s.d. = 459 g) and Chinstrap (mean = 3,733 g, s.d. = 384 g) did not diﬀer significantly, while Gentoo
penguins (mean = 5,076 g, s.d. = 504 g) diﬀered significantly from both species.

F(2,339): df of species, residuals