---
title: "Week 6 Assignment"
author: "Jessalyn Chuang"
output: pdf_document
geometry: margin=2.54cm
---

```{r}
# Load necessary packages.
library(here)
library(tidyverse)
library(moments)
library(dplyr)
```

#(1) Coastal Giant Salamander-
a. Filter the dataset only for coastal giant salamanders in cascades (C), pools (P), and side
channels (SC). Create a figure that helps you to evaluate whether their snout-to-tail length 
(length_2_mm) by habitat type (unittype) is normally distributed. You may also calculate skew
and kurtosis values to help with your decision-making. Are these data normally distributed?
Why or why not? If they are not, apply a log-transform (log10()) to the length data and
re-evaluate if the data now meets the criteria to be considered normally-distributed.

```{r}
vertebrates_data <- read.csv(file = here("./Assignments/week6/and_vertebrates.csv"),
                           stringsAsFactors = TRUE)

salamander_data <- filter(vertebrates_data, 
                              species == "Coastal giant salamander",
                              unittype %in% c('C', 'P', 'SC'))

#Rice's rule to find number of columns
n_C <- sum(salamander_data$unittype == 'C')
n_P <- sum(salamander_data$unittype == 'P')
n_SC <- sum(salamander_data$unittype == 'SC')
range_C <- salamander_data %>%
  filter(unittype == 'C') %>%
  summarize(range = max(length_2_mm) - min(length_2_mm)) %>%
  pull(range)
range_P <- salamander_data %>%
  filter(unittype == 'P') %>%
  summarize(range = max(length_2_mm) - min(length_2_mm)) %>%
  pull(range)
range_SC <- salamander_data %>%
  filter(unittype == 'SC') %>%
  summarize(range = max(length_2_mm) - min(length_2_mm)) %>%
  pull(range)

bin_number_C <- round(2*n_C^(1/3))
bin_number_P <- round(2*n_P^(1/3))
bin_number_SC <- round(2*n_SC^(1/3))

bin_width_C <- range_C/(2*n_C^(1/3))
bin_width_P <- range_P/(2*n_P^(1/3))
bin_width_SC <- range_SC/(2*n_SC^(1/3))

#check if we should do rice's rule or just let it choose like what was done in lab?

salamander_fig1 <- ggplot(salamander_data, aes(x = length_2_mm, fill = unittype)) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange", "pink", "purple")) +
  labs(x = "Snout-to-Tail Length(mm)", y = "Count") +
  facet_grid(.~unittype) +
  theme_bw() +
  theme(legend.position = "none")

salamander_fig1
```

```{r}
# Separate data by unittype
C_dat <- salamander_data %>%
filter(unittype == "C")

P_dat <- salamander_data %>%
filter(unittype == "P")

SC_dat <- salamander_data %>%
filter(unittype == "SC")

# Calculate skew.
skewness(C_dat$length_2_mm, na.rm = TRUE)
skewness(P_dat$length_2_mm, na.rm = TRUE)
skewness(SC_dat$length_2_mm, na.rm = TRUE)

# Calculate kurtosis.
kurtosis(C_dat$length_2_mm, na.rm = TRUE)
kurtosis(P_dat$length_2_mm, na.rm = TRUE)
kurtosis(SC_dat$length_2_mm, na.rm = TRUE)
```

Skew results:
Cascades: 0.94
Pool: 1.11
Side Channels: 0.99

Kurtosis results:
Cascades: 4.69
Pool: 4.89
Side Channels: 4.94

Visually, the three data from the three habitats all seem to be positive skewed, with the majority of points concentrated at lower values. From the skew results, all three have values that suggest positive skewing as they are all greater than 0. Lastly for kurtosis, all three have values greater than 3, suggesting that the data is leptokurtic. Therefore, this data is not normally distributed!

```{r}
#log-transformation of data
salamander_data_transformed <- mutate(salamander_data, length_2_mm = log10(length_2_mm))
```

```{r}
salamander_fig2 <- ggplot(salamander_data_transformed, aes(x = length_2_mm, fill = unittype)) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange", "pink", "purple")) +
  labs(x = "Snout-to-Tail Length(mm)", y = "Count") +
  facet_grid(.~unittype) +
  theme_bw() +
  theme(legend.position = "none")

salamander_fig2
```

```{r}
# Separate data by unittype
C_dat_trans <- salamander_data_transformed %>%
filter(unittype == "C")

P_dat_trans <- salamander_data_transformed %>%
filter(unittype == "P")

SC_dat_trans <- salamander_data_transformed %>%
filter(unittype == "SC")

# Calculate skew.
skewness(C_dat_trans$length_2_mm, na.rm = TRUE)
skewness(P_dat_trans$length_2_mm, na.rm = TRUE)
skewness(SC_dat_trans$length_2_mm, na.rm = TRUE)

# Calculate kurtosis.
kurtosis(C_dat_trans$length_2_mm, na.rm = TRUE)
kurtosis(P_dat_trans$length_2_mm, na.rm = TRUE)
kurtosis(SC_dat_trans$length_2_mm, na.rm = TRUE)
```

Now, all three habitat types appear to be normally distributed from the visual inspection and from their skewness values all being around 0 and kurtosis values being around 3.

b. If you found the data from part a were not normally distributed and a log-transform did not
change this finding, you may stop here and proceed to question 2. If you found the data from
part a were normally distributed, conduct a Bartlett’s test for equal variance to determine if
these data also satisfy the need for homogeneity of variances across groups. Do these data
have approximately equal variances? Why or why not? Remember, the data may not pass
the Bartlett’s test, but if they adhere to the rule of thumb mentioned above, you may proceed
with a one-way ANOVA.

```{r}
# Perform Bartlett test.
salamander_var <- bartlett.test(salamander_data_transformed$length_2_mm, 
                                salamander_data_transformed$unittype)
# Examine results.
salamander_var
```

Since p-value is less than 0.05, I reject the null hypothesis stating that the variance between all groups included in
the data are the same in favor of the alternate hypothesis that states that the variance between all groups included in the data are not the same. Thus, this data does not pass the Bartlett's test. However, if the largest sample variance is less than 4 times the smallest sample variance, we may still assume variances are equal across samples and conduct an ANOVA. Checking this:

```{r}
C_var <- var(C_dat_trans$length_2_mm, na.rm = TRUE)
P_var <- var(P_dat_trans$length_2_mm, na.rm = TRUE)
SC_var <- var(SC_dat_trans$length_2_mm, na.rm = TRUE)

C_var
P_var
SC_var
```

Variance results:
Cascades: 0.02
Pool: 0.02
Side Channels: 0.02

When finding their variances, they actually all turned out equal, so we may assume variances are approximately equal across samples and conduct an ANOVA.

c. If you found the data from part b did not display approximately equal variances, you may
stop here. If you found the data from part b did display approximately equal variances, conduct
a one-way ANOVA to see if there is a significant diﬀerence between coastal giant salamander
lengths across cascades, pools, and side channels of Mack Creek. If you find evidence of
significant diﬀerences, perform a post-hoc Tukey’s HSD test to determine which are significant
from which habitats. Communicate your findings as a figure (with an appropriate caption)
and in a sentence, as it might appear in a final report.

```{r}
salamander_ANOVA <- aov(length_2_mm ~ unittype, data = salamander_data_transformed)
summary(salamander_ANOVA)

#Perform a post-hoc Tukey's HSD test since p < 0.001
salamander_Tukey <- TukeyHSD(salamander_ANOVA)
salamander_Tukey
```

```{r}
salamander_summary <- salamander_data %>%
  group_by(unittype) %>%
  summarize(mean = mean(length_2_mm, na.rm = TRUE),
  sd = sd(length_2_mm, na.rm = TRUE)) %>%
  ungroup()

salamander_summary
```

Salamander species displayed significant diﬀerences in Snout-to-Tail length as determined by one-way ANOVA (F(2, 11411) = 343.6, p < 0.001). Post-hoc testing by Tukey’s HSD revealed that mean Snout-to-Tail length for salamanders in the Cascades habitat (mean =
104 mm, s.d. = 34.7 mm), Pool habitat (mean = 101 mm, s.d. = 36.1 mm), and Side Channels habitat (mean = 93.4 mm, s.d. = 30.5 mm) all differed significantly from each other.

#(2) Cutthroat Trout-

a. Filter the dataset only for cutthroat trout and create a figure that helps you to evaluate
whether their snout-to-fork length (length_1_mm) by reach (reach) is normally distributed. You
may also calculate skew and kurtosis values to help with your decision-making. Are these data
normally distributed? Why or why not? If they are not, apply a log-transform (log10()) to
the length data and re-evaluate if the data now meets the criteria to be considered normally-
distributed.

```{r}
trout_data <- filter(vertebrates_data, 
                              species == "Cutthroat trout")

trout_fig1 <- ggplot(trout_data, aes(x = length_1_mm, fill = reach)) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange", "pink", "purple")) +
  labs(x = "Snout-to-Fork Length (mm)", y = "Count") +
  facet_grid(.~reach) +
  theme_bw() +
  theme(legend.position = "none")

trout_fig1
```

```{r}
# Separate data by unittype
L_dat <- trout_data %>%
filter(reach == "L")

M_dat <- trout_data %>%
filter(reach == "M")

U_dat <- trout_data %>%
filter(reach == "U")

# Calculate skew.
skewness(L_dat$length_1_mm, na.rm = TRUE)
skewness(M_dat$length_1_mm, na.rm = TRUE)
skewness(U_dat$length_1_mm, na.rm = TRUE)

# Calculate kurtosis.
kurtosis(L_dat$length_1_mm, na.rm = TRUE)
kurtosis(M_dat$length_1_mm, na.rm = TRUE)
kurtosis(U_dat$length_1_mm, na.rm = TRUE)
```

Skew results:
L: 0.40
M: 0.26
U: 0.38

Kurtosis results:
L: 2.28
M: 2.16
U: 2.10

Visually, the three data sets appear bi-modal. From the skew results, all three have values slightly above 0 indicating slight positive skewing. All three have kurtosis values less than 3, indicating that they may be platykurtic. Overall, they do not seem normally distributed.

```{r}
#log-transformation of data
trout_data_transformed <- mutate(trout_data, length_1_mm = log10(length_1_mm))
```

```{r}
trout_fig2 <- ggplot(trout_data_transformed, aes(x = length_1_mm, fill = reach)) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange", "pink", "purple")) +
  labs(x = "Snout-to-Fork Length (mm)", y = "Count") +
  facet_grid(.~reach) +
  theme_bw() +
  theme(legend.position = "none")

trout_fig2
```

```{r}
# Separate data by unittype
L_dat_trans <- trout_data_transformed %>%
filter(reach == "L")

M_dat_trans <- trout_data_transformed %>%
filter(reach == "M")

U_dat_trans <- trout_data_transformed %>%
filter(reach == "U")

# Calculate skew.
skewness(L_dat_trans$length_1_mm, na.rm = TRUE)
skewness(M_dat_trans$length_1_mm, na.rm = TRUE)
skewness(U_dat_trans$length_1_mm, na.rm = TRUE)

# Calculate kurtosis.
kurtosis(L_dat_trans$length_1_mm, na.rm = TRUE)
kurtosis(M_dat_trans$length_1_mm, na.rm = TRUE)
kurtosis(U_dat_trans$length_1_mm, na.rm = TRUE)
```

After applying the log 10 transformation, the data is still bi-modal as apparent from a visual inspection. While skewness was brought closer to 0, kurtosis values are still less than 3, and so these distributions are considered platykurtic. The data are still not normally distributed.

b. If you found the data from part a were not normally distributed and a log-transform did
not change this finding, you may stop here. If you found the data from part a were normally
distributed, conduct a Bartlett’s test for equal variance to determine if these data also satisfy
the need for homogeneity of variances across groups. Do these data have approximately similar
variances? Why or why not? Remember, the data may not pass the Bartlett’s test, but if
they adhere to the rule of thumb mentioned above, you may proceed with a one-way ANOVA.

c. If you found the data from part b did not display equal variances, you may stop here. If
you found the data from part b did display equal variances, conduct a one-way ANOVA to see
if there is a significant diﬀerence between cutthroat trout lengths across lower, middle, and
upper reaches of Mack Creek. If you find evidence of significant diﬀerences, perform a post-
hoc Tukey’s HSD test to determine which are significant from which reaches. Communicate
your findings as a figure (with an appropriate caption) and a sentence, as it might appear in
a final report.