---
title: "Week 9 Assignment"
author: "Jessalyn Chuang"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
---

#(1) Logistic Regression
Download the dataset jrn_lizard.csv from Canvas. This dataset contains data on lizards
captured using pitfall traps by researchers at the Jornada Basin Long Term Ecological Research
site in southern New Mexico. For this portion of the assignment, you must build a logistic
regression model to investigate the following research question:

Do snout-to-vent length (SV_length), sex (sex), and vegetation zone (zone) at time of capture
significantly predict if a lizard tail is recorded as whole (tail = W)?

Before performing this regression, filter the dataset only for side-blotched lizards (Uta stans-
buriana or spp = UTST). Also, for context, the four possible vegetation zones are black grama grassland (zone = G), creosotebush shrubland (zone = C), mesquite duneland (zone = M), and tarbush shrubland (zone = T).

Using the workflow in today’s lab as a guide, work through the steps of performing a logistic
regression using these data, including exploratory data visualizations. You should report the
results of your final model in text form (1-2 sentences) as well as provide 2 predictive figures
showing the change in probability of a lizard tail being whole as lizard length increases, with
one including lines colored by sex and another including lines colored by vegetation zone.

```{r}
#install.packages("naniar", dependencies = TRUE)
#install.packages("DHARMa")
#install.packages("gtsummary")

library(here)
library(tidyverse)
library(naniar)
library(DHARMa)
library(gtsummary)
library(broom)

lizard_data <- read_csv("./Assignments/week9/jrn_lizard.csv")
utst_lizards <- lizard_data %>% 
  filter(spp == "UTST") %>% 
  filter(zone %in% c('G','C','M','T')) %>% 
  select(SV_length, zone, sex, tail)
utst_lizards_clean <- na.omit(utst_lizards)
```

##Visualizing Data
```{r}
hist_SV_length <- ggplot(utst_lizards_clean, aes(x = SV_length)) +  
    geom_histogram() +  
    theme_bw() +
  ggtitle("Count of lizards for each Snout-to-Vent Length")
  
hist_SV_length

bar_sex <- ggplot(utst_lizards_clean, aes(x = sex)) +  
    geom_bar() +  
    theme_bw() +
  ggtitle("Count of lizards for each Sex")
bar_sex

bar_zone <- ggplot(utst_lizards_clean, aes(x = zone)) +  
    geom_bar() +  
    theme_bw() +
  ggtitle("Count of lizards in each Zone")
bar_zone

bar_tail <- ggplot(utst_lizards_clean, aes(x = tail)) +  
    geom_bar() +  
    theme_bw() + 
  ggtitle("Count of lizards for each Tail Type")
bar_tail
```

Observations from distributions: 
1)SV_length: concentrated between 25 and 55, some potential outliers (values above 70 and below 15).
2) Sex: three major categories with F being most common
3) Zone: Zone M (mesquite duneland) has the highest count
4) Tail: There are much more lizards recorded with a Whole tail. There is a strong class imbalance here.

Relationships between predictors (SV_length, Sex, Zone) and the outcome variable (Tail)
```{r}
boxplot_1 <- ggplot(utst_lizards_clean, aes(x = tail, y = SV_length)) +  
    geom_boxplot() +  
    theme_bw() +
  ggtitle("Range of Snout-to-Vent Length by Tail Type")
boxplot_1

boxplot_2 <- ggplot(utst_lizards_clean, aes(x = sex, y = SV_length)) +  
    geom_boxplot() +  
    theme_bw() +
  ggtitle("Range of Snout-to-Vent Length by Sex")
boxplot_2

boxplot_3 <- ggplot(utst_lizards_clean, aes(x = zone, y = SV_length)) +  
    geom_boxplot() +  
    theme_bw() + 
  ggtitle("Tail") +
  ggtitle("Range of Snout-to-Vent Length by Zone")
boxplot_3
```
These three box plots display the distribution of snout-to-vent (SV) length in lizards, grouped by tail type, sex, and zone. The first plot shows that lizards with B tend to have slightly longer SV lengths than those with whole tails, though the difference is small and both groups include a few outliers. The second plot reveals that juveniles (J) have the shortest SV lengths, while males (M) and females (F) show greater lengths with overlapping ranges:males have a slightly higher median than females. The third plot compares SV length by geographic zone, showing relatively similar distributions in zones C, G, and M, with zone T standing out as having the highest median and upper range of SV lengths. Overall, SV length appears to vary meaningfully by sex and zone, and to a lesser extent by tail condition.

##Fitting regression model
```{r}
#Set tail type column to be a factor with a reference level of B
utst_lizards_clean$tail <- factor(utst_lizards_clean$tail, levels = c("B", "W"))
utst_lizards_clean$sex <- factor(utst_lizards_clean$sex, levels = c("J", "F", "M"))
utst_lizards_clean$zone <- factor(utst_lizards_clean$zone, levels = c("C", "G", "M", "T"))

tail_type_mod <- glm(tail ~ SV_length + sex + zone, data = utst_lizards_clean,
                     family = "binomial")
```

##Evaluating model diagnostics
```{r}
summary(tail_type_mod)

simulateResiduals(tail_type_mod) %>% plot()
```

The residual diagnostics indicate a good fit of the logistic regression model. The QQ plot on the left shows that the observed residuals closely follow the expected uniform distribution, with no significant deviation (KS test: p = 0.255). The residual vs. predicted plot on the right shows that residuals are evenly scattered without clear patterns or trends, suggesting no major issues with heteroscedasticity. Overall, the residuals appear to follow what we should expect.

##Intepreting model and communicating results

The results of a logistic regression indicate that the log-odds of having a whole tail significantly decrease with increasing Snout-to-Vent length (B = -0.045, p < 0.001). Compared to lizards in the reference vegetation zone C, lizards in zone G (B = -0.565, p = 0.001) and zone T (B = -0.404, p = 0.025) have significantly lower log-odds of having a whole tail. There is no significant difference in tail status for lizards in zone M (p = 0.063), or between juveniles and either sex (sexF or sexM).

#2 Predictive Figures
showing the change in probability of a lizard tail being whole as lizard length increases, with
one including lines colored by sex and another including lines colored by vegetation zone
```{r}
SV_length_vector <- rep(seq(from = 0, to = 90), 3)

# Keep zone constant at "C"
zone_vector <- rep("C", length(SV_length_vector)) 
  
# Simulate sex data repeating "F", "J", "M"
sex_vector <- rep(c("F", "J", "M"), each = length(seq(from = 0, to = 90)))
  
data_pred <-  data.frame(SV_length_vector, zone_vector, sex_vector)
colnames(data_pred) <- c("SV_length", "zone", "sex")

prediction <- predict(tail_type_mod,
                      newdata = data_pred,
                      type = "response",
                      se.fit = TRUE)

data_fig <-  data.frame(data_pred,
                        prediction$fit,
                        prediction$se.fit)

# Rename columns.
colnames(data_fig) <- c("SV_length","zone", "sex", "probability", "se")

#Graph probability of a lizard tail being whole as lizard length increases, with
#lines colored by sex 
ggplot(data_fig, aes(x = SV_length, y = probability)) +
  geom_line(aes(color = sex)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = sex), alpha = 0.3) +
  labs(x = "SV Length",
       y = "Probability of Having Whole Tail",
       color = "Sex",
       fill = "Sex") +
  theme_bw() + 
  ggtitle("Probability of Whole Tail with Change in Lizard Length by Sex")
```

This figure shows the modeled probability of a lizard having a whole tail as a function of snout-to-vent length, with separate curves for females (F), juveniles (J), and males (M). Across all sex categories, the probability of having a whole tail decreases steadily as lizard length increases. While all three groups follow a similar downward trend, juveniles consistently show a slightly higher probability of a whole tail at larger lengths, whereas females show the lowest probabilities across much of the SV length range. However, the confidence intervals for each group (shaded areas) broadly overlap, alluding to differences between sexes not being statistically significant (which is what was found from the GLM results).

```{r}
SV_length_vector <- rep(seq(from = 0, to = 90), 4)

# Simulate zone data repeating "C", "G", "M", "T"
zone_vector <- rep(c("C", "G", "M", "T"), each = length(seq(from = 0, to = 90))) 
  
# Keep sex constant at "F"
sex_vector <- rep("J", length(SV_length_vector))
  
data_pred <-  data.frame(SV_length_vector, zone_vector, sex_vector)
colnames(data_pred) <- c("SV_length", "zone", "sex")

prediction <- predict(tail_type_mod,
                      newdata = data_pred,
                      type = "response",
                      se.fit = TRUE)

data_fig <-  data.frame(data_pred,
                        prediction$fit,
                        prediction$se.fit)

# Rename columns.
colnames(data_fig) <- c("SV_length","zone", "sex", "probability", "se")
#Graph probability of a lizard tail being whole as lizard length increases, with
#lines colored by zone 
ggplot(data_fig, aes(x = SV_length, y = probability)) +
  geom_line(aes(color = zone)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = zone), alpha = 0.3) +
  labs(x = "SV Length",
       y = "Probability of Having Whole Tail",
       color = "Zone",
       fill = "Zone") +
  theme_bw() + 
  ggtitle("Probability of Whole Tail with Change in Lizard Length by Zone")
```

This figure illustrates how the probability of having a whole tail changes with SV length across four vegetation zones (C = creosotebush shrubland, G = black grama grassland, M = mesquite duneland, T = tarbush shrubland). In all zones, the probability of whole tail declines as lizard length increases. Lizards in Zone C consistently show the highest probability of having a whole tail across the SV length range, while those in Zones G, M, and T have lower probabilities, with Zone G showing the steepest decline. Although the confidence intervals overlap, the separation between the curves indicates potential differences in tail loss patterns by zone.

#(2) Poisson Regression

Download the dataset jrn_lizard_npp.csv from Canvas. This dataset combines counts of lizards
captured using pitfall traps with percent cover of diﬀerent plant species at the Jornada Basin
Long Term Ecological Research. For this portion of the assignment, you must build a poisson
regression model to investigate the following research question:

Do season (season) and plant species percent cover (BOER, LATR, PRGL, SCBR) significantly predict
lizard counts (lizard_count)?

These data have already been filtered only for side-blotched lizards (Uta stansburiana). Sea-
sons include fall (F), winter (W), and summer (S). In addition, the four plant species whose
percent covers are reported here are creosote bush (Larrea tridentata or LATR), burrograss
(Scleropogon brevifolius or SCBR), honey mesquite (Prosopis glandulosa or PRGL), and black
grama grass (Bouteloua eriopoda or BOER). Note, “percent cover” refers to the percent of a
given quadrat, or sampling area, that a species occupies.

Using the workflow in today’s lab as a guide, work through the steps of performing a poisson
regression using these data, including exploratory data visualizations. You are not expected
to include all four plant species in your model structure but should use your initial data
visualizations to help determine which you feel are best suited to help predict lizard counts.
You should report the results of your final model in text form (1-2 sentences). If you choose
to remove outliers during the modeling process, be sure to report how many you removed and
your justification for doing so.

```{r}
lizard_npp <- read_csv("./Assignments/week9/jrn_lizard_npp.csv")
```

##Visualizing Data
```{r}
hist_BOER <- ggplot(lizard_npp, aes(x = BOER)) +  
    geom_histogram() +  
    theme_bw() + 
  ggtitle("Count of lizards at varying BOER levels")
hist_BOER

hist_LATR <- ggplot(lizard_npp, aes(x = LATR)) +  
    geom_histogram() +  
    theme_bw() +
  ggtitle("Count of lizards at varying LATR levels")
hist_LATR

hist_PRGL <- ggplot(lizard_npp, aes(x = PRGL)) +  
    geom_histogram() +  
    theme_bw() +
  ggtitle("Count of lizards at varying PRGL levels")
hist_PRGL

hist_SCBR <- ggplot(lizard_npp, aes(x = SCBR)) +  
    geom_histogram() +  
    theme_bw() +
  ggtitle("Count of lizards at varying SCBR levels")
hist_SCBR

bar_season <- ggplot(lizard_npp, aes(x = season)) +  
    geom_bar() +  
    theme_bw() +
  ggtitle("Count of lizards Collected in each Season")
bar_season
```

BOER, LATR, and PRGL all seem to have values concentrated on one end with a few number of outliers at the higher end. SCBR has a bit more counts spread out across percentages, though counts are more concentrated between 0-1%. Counts across seasons are fairly balanced, with Fall only seeing slightly more counts.

Investigating multi-collinearity:

```{r}
cor.test(lizard_npp$BOER, lizard_npp$LATR)
cor.test(lizard_npp$BOER, lizard_npp$PRGL)
cor.test(lizard_npp$BOER, lizard_npp$SCBR, use = "complete.obs")
cor.test(lizard_npp$LATR, lizard_npp$PRGL)
cor.test(lizard_npp$LATR, lizard_npp$SCBR, use = "complete.obs")
cor.test(lizard_npp$PRGL, lizard_npp$SCBR, use = "complete.obs")
```

After checking all the possible pairwise combinations of the species, it looks like LATR and SCBR are the only two species that are not highly correlated with each other (others have correlation values > 0.62) so I will only include these two species in my final model.

##Fitting regression model
```{r}
lizard_npp$season <- factor(lizard_npp$season, levels = c("F", "S", "W"))

lizard_count_mod <- glm(lizard_count ~ LATR + SCBR + season, data = lizard_npp,
                     family = "poisson")
```

##Evaluating model diagnostics
```{r}
summary(lizard_count_mod)

plot(lizard_count_mod)
```

Based on the plot of the residuals, point 47 is considered an outlier on the Residuals vs Fitted, Scale-Location, and Residuals vs Leverage plots. Point 1 is also considered an outlier on the Residuals vs Leverage plot. These two points are removed, and I refitted the model on the dataset without the outliers:

```{r}
#remove outliers using rownumbers
lizard_npp_outliers_removed <-  lizard_npp[-c(1, 47),]
lizard_count_mod_2 <- glm(lizard_count ~ LATR + SCBR + season, data = lizard_npp_outliers_removed,
                     family = "poisson")

summary(lizard_count_mod_2)

plot(lizard_count_mod_2)
```

The model still seems to have slightly non-normal residuals, uneven variance, and a few influential points. Point 48 consistently shows up as an extreme outlier across all three plots, so I will remove this last point and refit the model:

```{r}
#remove outliers using rownumbers
lizard_npp_outliers_removed_2 <-  lizard_npp[-c(48),]
lizard_count_mod_3 <- glm(lizard_count ~ LATR + SCBR + season, data = lizard_npp_outliers_removed_2,
                     family = "poisson")

summary(lizard_count_mod_3)

plot(lizard_count_mod_3)
```

The residual plots overall look better. There are now 2 points considered outliers on the Residuals vs Leverage plot, but I am opting to keep those as i've already removed three observations, and the other three plots look more stable with residual residuals appearing more homoscedastic and normally distributed than before.

##Intepreting model and communicating results

The results of a poisson regression suggest that both the winter season (B = -0.027, p = 0.62) and percent cover of creosote bush (B = -0.042, p = 0.20) do not have a strong effect on lizard counts.  However, the summer season does have a significant effect (B = -0.817, p <0.001), and so does the percent cover of burrograss (B = 0.437, p < 0.001). I removed observations 1, 47, and 48 since they appeared as outliers in
the residual plots and may have skewed data results inaccurately to show higher correlations when in reality they are weaker. Note, these coeﬃcients are for estimation of log-transformed counts.

#(3) GitHub customization
Having created a repository that functions as your homepage on GitHub, go ahead and cus-
tomize it in whatever way you think is most helpful for you and for future collaborators to
see. At a minimum, you must include at least one image (it does not need to be a photo of
you, but it should be properly attributed if you did not take it), a few sentences of text, and
a hyperlink to another site (e.g., LinkedIn). For this assignment, all you have to do is include
a hyperlink to your GitHub profile

My GitHub profile: https://github.com/jessalynlc/jessalynlc
