---
title: "week4assignment_JC"
output: pdf_document
date: "2025-02-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(here)
library(dplyr)
library(moments)
library(vcdExtra)
```

#(1) Data Tidying and Visualization

##a. Tidying 

```{r }
urchins_data <- read.csv(file = here("./Assignments/week4/urchins.csv"),
                           stringsAsFactors = TRUE)   

urchins_data <- urchins_data[,c("YEAR", "MONTH", "SITE", "TREATMENT", "SIZE", 
                                "COUNT", "COMMON_NAME")]

urchins_data <- filter(urchins_data, SITE == "MOHK" & (YEAR == 2015 | YEAR == 2016))

urchins_data <- expand.dft(urchins_data, freq = "COUNT")
```

##b. Histogram of size distributions of red urchins in 2015 and 2016

ASK ABOUT NUM OF BINS HERE?

To find the number of bins, use Rice's rule:
In 2015, n = 773: 2*773^(1/3) = 18.35
In 2016, n = 400: 2*400^(1/3) = 14.74

Average between the two is number of bins: 16.5 --> 17

```{r}
ggplot(urchins_data, aes(x = SIZE, fill = TREATMENT)) +
  geom_histogram(bins = 17) +   
  facet_wrap(~ YEAR) +
  labs(title = "Size Distribution of Red Urchins (2015-2016)",
       x = "Urchin Size (mm)",
       y = "Count",
       fill = "Treatment",
       caption = "Figure 1: Histogram of red urchin size distributions at Mohawk Reef in 2015 and 2016,
       categorized by treatment type. The x-axis represents urchin size (mm), and the y-axis indicates 
       the count of individuals in each size bin.") +
  theme(plot.caption = element_text(hjust = 0.5)) 
```

#(2) Confidence Intervals

##a. 95% two-sided confidence interval for red urchin sizes in 2015 across treatment types

```{r}
#2015
urchins_2015 <- urchins_data %>%
  filter(YEAR == 2015 & COMMON_NAME == "Red Urchin") 
mean_2015 <- mean(urchins_2015$SIZE)
urchins_2015_CI <- t.test(urchins_2015$SIZE, mu = mean_2015, conf.level = 0.95, 
                          alternative = "two.sided")
print(paste("The 95% confidence interval for red urchin sizes in 2015 is [", 
            round(urchins_2015_CI$conf.int[1], 2), ",", 
            round(urchins_2015_CI$conf.int[2], 2), "]"))

#2016
urchins_2016 <- urchins_data %>%
  filter(YEAR == 2016 & COMMON_NAME == "Purple Urchin") 
mean_2016 <- mean(urchins_2016$SIZE)
urchins_2016_CI <- t.test(urchins_2016$SIZE, mu = mean_2016, conf.level = 0.95, 
                          alternative = "two.sided")
print(paste("The 95% confidence interval for purple urchin sizes in 2016 is [", 
            round(urchins_2016_CI$conf.int[1], 2), ",", 
            round(urchins_2016_CI$conf.int[2], 2), "]"))
```

#(3) One-sample Hypothesis Tests

##(a) One-sided one-sample t-test to see if red urchins in 2015 from control group are significantly less than the mean red urchin size of 9 cm reported by researchers at another location

```{r}
urchins_2015_control <- urchins_data %>%
  filter(YEAR == 2015 & COMMON_NAME == "Red Urchin" & TREATMENT == "CONTROL") 
urchins_2015_control_test <- t.test(urchins_2015_control$SIZE, mu = 9, 
                                    conf.level = 0.95, alternative = "less")
print(paste("Resulting p-value = ", urchins_2015_control_test$p.value))
```

With a 95% confidence leval, since p-value is less than 0.05, we reject the null hypothesis in favor of the alternative hypothesis which is that red urchins in the 2015 from the control group have significantly lower sizs that the mean red urch size of 9 cm reported by researchers at another location.

#(4) Two-sample Hypothesis Tests

##(ai) Two-sample two-sided t-test for mean red urchin sizes between the two treatments in 2015

```{r}
red_2015_control <- urchins_2015_control
red_2015_annual <- urchins_data %>%
  filter(YEAR == 2015 & COMMON_NAME == "Red Urchin" & TREATMENT == "ANNUAL REMOVAL") 

#testing for equal variances first
red_2015_vartest <- var.test(x = red_2015_control$SIZE, y = red_2015_annual$SIZE,
                             alternative = "two.sided",
                             conf.level = 0.95)

red_2015_vartest

#The p-value <0.05, so we reject the hypothesis and we retain the alternative
#hypothesis that the variances are not equal

red_2015_ttest <- t.test(red_2015_control$SIZE, red_2015_annual$SIZE)

red_2015_ttest
```

The mean red urchin size in 2015 was not significantly different between the control group (mean ± standard deviation = 8.18 ± 1.38) and the annual removal group (mean ± standard deviation = 8.22 ± 0.92) as determined by a two-sample, two-sided t-test (t = -0.29, df = 313.51, p = 0.77, α = 0.05)

##(aii) Two-sample two-sided t-test for mean red urchin sizes between the two treatments in 2016

```{r}
red_2016_control <- urchins_data %>%
  filter(YEAR == 2016 & COMMON_NAME == "Red Urchin" & TREATMENT == "CONTROL") 
red_2016_annual <- urchins_data %>%
  filter(YEAR == 2016 & COMMON_NAME == "Red Urchin" & TREATMENT == "ANNUAL REMOVAL") 

#testing for equal variances first
red_2016_vartest <- var.test(x = red_2016_control$SIZE, y = red_2016_annual$SIZE,
                             alternative = "two.sided",
                             conf.level = 0.95)

red_2016_vartest

#The p-value <0.05, so we reject the hypothesis and we retain the alternative
#hypothesis that the variances are not equal

red_2016_ttest <- t.test(red_2016_control$SIZE, red_2016_annual$SIZE)

red_2016_ttest
```

The mean red urchin size in 2016 was significantly different between the control group (mean ± standard deviation = 8.55 ± 1.06) and the annual removal group (mean ± standard deviation = 8.01 ± 0.85) as determined by a two-sample, two-sided t-test (t = 4.0132, df = 190.00, p = 8.597e-05, α = 0.05)

##(b) Two-sample two-sided t-test for mean red urchin sizes in the control group between 2015 and 2016

```{r}

#testing for equal variances first
red_control_vartest <- var.test(x = red_2015_control$SIZE, y = red_2016_control$SIZE,
                             alternative = "two.sided",
                             conf.level = 0.95)

red_control_vartest

#The p-value <0.05, so we reject the hypothesis and we retain the alternative
#hypothesis that the variances are not equal

red_control_ttest <- t.test(red_2015_control$SIZE, red_2016_control$SIZE)

red_control_ttest
```

The mean red urchin size was significantly different between the 2015 data (mean ± standard deviation = 8.18 ± 1.38) and the 2016 data (mean ± standard deviation = 8.55 ± 1.06) as determined by a two-sample, two-sided t-test (t = -2.54, df = 252.7, p = 0.01, α = 0.05)