---
title: "Week 7 Assignment"
author: "Jessalyn Chuang"
output: pdf_document
geometry: margin=2.54cm
---

```{r}
# Load necessary packages.
library(here)
library(tidyverse)
#install.packages('dunn.test')
library(dunn.test)
library(gt)
#install.packages('paletteer')
library(paletteer)
#install.packages("tibble")
library(tibble)
```

#(1) Mann-Whitney U and Wilcoxon Signed-Rank Tests

a. Filter the dataset only for cutthroat trout. Create a figure that helps you to demonstrate
why their snout-to-fork length (length_1_mm) by forest treatment (section) is not normally
distributed. Clear-cut regions of the forest are denoted with CC, and old-growth regions of
the forest are denoted as OG. Run a Mann-Whitney U test to see if ranks of fish lengths are
significantly diﬀerent between forest treatment types. Write out your findings as you would in
a report, including measures of central tendency and data spread for both forest treatments.

```{r}
vertebrates_data <- read.csv(file = here("./Assignments/week7/and_vertebrates.csv"),
                           stringsAsFactors = TRUE)

trout_data <- filter(vertebrates_data, 
                              species == "Cutthroat trout")

trout_fig1 <- ggplot(trout_data, aes(x = length_1_mm, fill = section)) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange", "pink", "purple")) +
  labs(x = "Snout-to-Fork Length (mm)", y = "Count") +
  facet_grid(.~section) +
  theme_bw() +
  theme(legend.position = "none")

trout_fig1

# Save data by section type.
trout_CC <- trout_data %>%
filter(section == "CC")
trout_OG <- trout_data %>%
filter(section == "OG")

mann_test <- wilcox.test(trout_CC$length_1_mm, trout_OG$length_1_mm,
paired = FALSE)

mann_test

trout_CC_mean <- mean(trout_CC$length_1_mm, na.rm = TRUE)
trout_CC_sd <- sd(trout_CC$length_1_mm, na.rm = TRUE)

trout_OG_mean <- mean(trout_OG$length_1_mm, na.rm = TRUE)
trout_OG_sd <- sd(trout_OG$length_1_mm, na.rm = TRUE)
```

Cutthroat trout species displayed significant diﬀerences in Snout-to-Fork length as determined by a Mann-Whitney U-Test (W = 55178843, p < 0.001). Therefore, we accept the alternative hypothesis that the ranks of the Snout-to-Fork length for Cutthroat trout in Clear-Cut regions (mean = 85.31 mm, s.d. = 36.03 mm) and Old-Growth regions (mean = 981.43 mm, s.d. = 34.46 mm) are significantly different.

b. Navigate to Canvas and download the trout_recapture.csv dataset; these data are originally
part of the and_vertebrates.csv dataset, but they have been filtered to include only trout that
were captured and tagged in 2014 and then re-captured in 2015. Create a figure that helps you
demonstrate why their snout-to-fork length (length_1_mm) by year is not normally distributed.
Run a Wilcoxon Signed-Rank test to see if ranks of fish lengths are significantly diﬀerent
between years. Write out your findings as you would in a report, including measures of
central tendency and data spread for both years.

```{r}
trout_recapture <- read.csv(file = here("./Assignments/week7/trout_recapture_v2.csv"),
                           stringsAsFactors = TRUE)

trout_fig2 <- ggplot(trout_recapture, aes(x = length_1_mm, fill = factor(year))) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange", "pink")) +
  labs(x = "Snout-to-Fork Length (mm)", y = "Count") +
  facet_grid(.~year) +
  theme_bw() +
  theme(legend.position = "none")

trout_fig2

# Save data by year.
trout_2014 <- trout_recapture %>%
filter(year == 2014)
trout_2015 <- trout_recapture %>%
filter(year == 2015)

wilcox_test <- wilcox.test(trout_2014$length_1_mm, trout_2015$length_1_mm,
paired = TRUE)

wilcox_test

trout_2014_mean <- mean(trout_2014$length_1_mm, na.rm = TRUE)
trout_2014_sd <- sd(trout_2014$length_1_mm, na.rm = TRUE)

trout_2015_mean <- mean(trout_2015$length_1_mm, na.rm = TRUE)
trout_2015_sd <- sd(trout_2015$length_1_mm, na.rm = TRUE)
```

Recaptured Cutthroat trout displayed significant diﬀerences in Snout-to-Fork length as determined by a Wilcoxon Signed-Rank Test (V = 625.5, p < 0.001). Therefore, we accept the alternative hypothesis that the ranks of the Snout-to-Fork length for Cutthroat trout in 2014 (mean = 105.48 mm, s.d. = 17.27 mm) and 2015 (mean = 119.63 mm, s.d. = 14.84 mm) are significantly different.

#(2) Kruskal Wallis and post-hoc Dunn’s Tests

a. Using the dataset you created above including only cutthroat trout, create a figure that
helps you to demonstrate why their snout-to-fork length (length_1_mm) by reach (reach) is
not normally distributed. Having verified this, run a Kruskal Wallis test to see if ranks of
fish lengths are significantly diﬀerent among stream reaches. If you find there is a significant
diﬀerence between at least two reaches, run a Dunn’s test. Write out your findings as you
would in a report, including a measure of central tendency and data spread for all three reach
sections.

```{r}
trout_fig3 <- ggplot(trout_data, aes(x = length_1_mm, fill = reach)) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange", "pink", "purple")) +
  labs(x = "Snout-to-Fork Length (mm)", y = "Count") +
  facet_grid(.~reach) +
  theme_bw() +
  theme(legend.position = "none")

trout_fig3

kruskal_test <- kruskal.test(length_1_mm ~ reach, data = trout_data)

kruskal_test

# There is a significant difference between at least two reaches
dunn_test <- dunn.test(trout_data$length_1_mm, trout_data$reach)

# Save data by reach.
trout_L <- trout_data %>%
filter(reach == "L")
trout_M <- trout_data %>%
filter(reach == "M")
trout_U <- trout_data %>%
filter(reach == "U")

trout_L_mean <- mean(trout_L$length_1_mm, na.rm = TRUE)
trout_L_sd <- sd(trout_L$length_1_mm, na.rm = TRUE)

trout_M_mean <- mean(trout_M$length_1_mm, na.rm = TRUE)
trout_M_sd <- sd(trout_M$length_1_mm, na.rm = TRUE)

trout_U_mean <- mean(trout_U$length_1_mm, na.rm = TRUE)
trout_U_sd <- sd(trout_U$length_1_mm, na.rm = TRUE)
```

Cutthroat trout species displayed significant diﬀerences in Snout-to-Fork length as determined by a Kruskall-Wallis Rank Sum Test (χ²(2) = 77.394, p < 0.001). Post-hoc testing by Dunn's Test revealed that ranks of fish lengths are significantly different between L reach (mean = 81.60 mm, s.d. = 34.71 mm) and M reach (mean = 87.01 mm, s.d. = 36.09 mm), and between M reach and U reach (mean = 82.19 mm, s.d. = 35.11 mm). There is no significant difference between L and U reaches.

b. Create a figure with an appropriate caption to display fish lengths by reach, including
notation to indicate the results of the Kruskal Wallis and Dunn’s tests (letters, etc.).

```{r}
letters_df <- tibble(
  reach = c("L", "M", "U"),
  Letter = c("a", "b", "a") # Corrected assignment
)

trout_summary <- trout_data %>%
  group_by(reach) %>%
  summarise(mean_length = mean(length_1_mm, na.rm = TRUE),
            sd_length = sd(length_1_mm, na.rm = TRUE),
            max_length = max(length_1_mm, na.rm = TRUE)) %>%
  left_join(letters_df, by = "reach")

# Create a boxplot with significance letters
trout_fig4 <- ggplot(trout_data, aes(x = reach, y = length_1_mm, fill = reach)) +
  geom_boxplot() +
  geom_text(data = trout_summary, aes(x = reach, y = max_length + 10, label = Letter), 
            size = 6, vjust = 0) +  # Position letters correctly
  scale_fill_manual(values = c("darkorange", "pink", "purple")) +
  labs(x = "Reach", y = "Snout-to-Fork Length (mm)", fill = "Reach",
       title = "Fish Length by Reach") +
  theme_bw() +
  theme(legend.position = "none")

# Display the final plot
print(trout_fig4)
cat("This boxplot shows the distribution of fish length across different reaches (L, M, U). The letters above each box indicate statistical groupings based on Dunn’s post-hoc test. Groups sharing the same letter are not significantly different, whereas groups with different letters indicate a significant difference in fish length (α = 0.05).")
```

#(3) Contingency Tables and the Chi-Squared Test

a. Using the dataset you created above including only cutthroat trout, create a contingency
table that includes counts of trout found in diﬀerent channel habitat types (unittype) by reach
(reach). If you’re unsure how to create this kind of table, think back to the group_by() and
summarize() functions you learned last week. Also, you may want to explore the pivot_wider()
and pivot_longer() functions for pivoting data from long to wide format. Create a table using
the gt() function with an appropriate caption to display raw fish counts.

```{r}
trout_summary_unittypebyreach <- trout_data %>%
  group_by(unittype, reach) %>%
  summarize(count = n(), .groups = "drop")

habitatbyreach_table <- trout_summary_unittypebyreach %>%
  select(unittype, reach, count) %>% 
  pivot_wider(names_from = reach, values_from = count)

# Make this table a dataframe
habitatbyreach_df <- as.data.frame(habitatbyreach_table)

# Add row totals
habitatbyreach_df <- habitatbyreach_df %>%
  mutate(Total = rowSums(select(., L, M, U), na.rm = TRUE))

habitat_gt <- habitatbyreach_df %>%
  gt() %>%
  tab_header(title = "Fish Habitat Counts by Reach") %>% 
  cols_label(
    unittype = "Habitat Type",
    L = "L Reach",
    M = "M Reach",
    U = "U Reach",
    Total = "Total" 
  ) %>%
  grand_summary_rows(
    columns = c(L, M, U, Total),  
    fns = list(Total = ~sum(., na.rm = TRUE)) 
  ) %>%
  data_color(
    columns = c(L, M, U),
    palette = c("#C0D8F0", "#5A7ECB"),
    alpha = 0.75
  )

# Display the table
habitat_gt

```

b. Before proceeding further, write out your null and alternative hypotheses for performing a
chi-squared test on this dataset. Then, run a chi-squared test on the counts dataset you have
created. You may want to use the column_to_rownames() function to make sure the only values
in cells of your contingency table are trout counts before running your chi-squared test. Write
out your findings as you would in a report, including language about the habitat type where
most fish are counted in each reach.

Null hypothesis: There is no significant association between habitat type and reach for Cutthroat trout.

Alternative hypothesis: There is a significant association between habitat type and reach for Cutthroat trout.

```{r}
habitat_matrix <- habitatbyreach_df %>%
  filter(!is.na(unittype)) %>%  # Remove NA rows in unittype
  column_to_rownames(var = "unittype") %>%  # Convert unittype to row names
  select(L, M, U) %>%  # Keep only numeric columns
  replace(is.na(.), 0)  # Replace NA counts with 0

# Run Chi-Square test
chi_test <- chisq.test(habitat_matrix)

chi_test
```

There is a significant association between habitat and reach for Cutthroat trout (χ²(12) = 1914.4, p<0.001).
